---
title: "Inspect vars info over time - Defunciones Fetales"
output:
  html_notebook:
    code_folding: hide
---

```{r}
vars_info_defun_fetal <- tar_read(vars_info_defun_fetal)
```

```{r results="asis"}
defun_fetal_raw <- tar_read(defun_fetal_raw)
```

Mirrors `inspect_vars_info.Rmd` (written for Nacimientos), but for
Defunciones Fetales (fetal deaths). To quickly visuallize what has changed,
we'll take a look at a few plots that show i) which variables exist in
which years and ii) how are those variables encoded (class, type and
unique values). This is an input to individually decide what to do with
each variable (e.g. include/exclude, how to homologate them when
categories have changed). See also `inspect_ddi_vars_defun_fetal.Rmd` for
the equivalent analysis based on the DDI codebooks rather than the raw
ingested data.

# Data as Ingested

```{r}
plotly::ggplotly(
  vars_info_defun_fetal |>
    mutate(value = 1) |>
    ggplot(aes(x = year, y = var_names, fill = value, text = glue::glue(
      "\n",
      "var_class: {var_class}",
      "var_type: {var_type}",
      "var_unique: {var_unique}",
      .sep = "\n"
    ))) +
    geom_tile() +
    theme(legend.position = "none")
) |>
  plotly::layout(
    legend = list(orientation = "h"), xaxis = list(title = NA, tickangle = -45)
  )
```

```{r}
plotly::ggplotly(
  vars_info_defun_fetal |>
    ggplot(aes(x = year, y = var_names, fill = var_class, text = glue::glue(
      "\n",
      "var_class: {var_class}",
      "var_type: {var_type}",
      "var_unique: {var_unique}",
      .sep = "\n"
    ))) +
    geom_tile()
) |>
  plotly::layout(
    legend = list(orientation = "h"), xaxis = list(title = NA, tickangle = -45)
  )
```

```{r}
plotly::ggplotly(
  vars_info_defun_fetal |>
    ggplot(aes(x = year, y = var_names, fill = var_type, text = glue::glue(
      "\n",
      "var_class: {var_class}",
      "var_type: {var_type}",
      "var_unique: {var_unique}",
      .sep = "\n"
    ))) +
    geom_tile()
) |>
  plotly::layout(
    legend = list(orientation = "h"), xaxis = list(title = NA, tickangle = -45)
  )
```

```{r}
plotly::ggplotly(
  vars_info_defun_fetal |>
    ggplot(aes(x = year, y = var_names, fill = var_unique, text = glue::glue(
      "\n",
      "var_class: {var_class}",
      "var_type: {var_type}",
      "var_unique: {var_unique}",
      .sep = "\n"
    ))) +
    geom_tile() +
    theme(legend.position = "none")
) |>
  plotly::layout(
    legend = list(orientation = "h"), xaxis = list(title = NA, tickangle = -45)
  )
```

# Checking vars

Nothing has been individually reviewed yet for this dataset (unlike
Nacimientos' `inspect_vars_info.Rmd`, which documents, variable by
variable, the homologation decisions behind `homologate_vars()`). The list
below is meant as the starting point for that same exercise here: for each
variable, look at the plots above (and at `inspect_ddi_vars_defun_fetal.Rmd`)
and decide whether/how to homologate it, then move its name from
"pending" to a `## variable_name` subsection documenting the decision,
mirroring `inspect_vars_info.Rmd`.

```{r}
already_checked <- c()
```

## Vars pending screening

```{r}
vars_info_defun_fetal$var_names |>
  unique() |>
  setdiff(already_checked) |>
  sort()
```
