---
title: "Inspect DDI variables over time - Defunciones No Fetales"
output:
  html_notebook:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r}
ddi_defun_nofetal_vars <- tar_read(ddi_defun_nofetal_vars)
```

While `inspect_vars_info_defun_nofetal.Rmd` looks at the data *as ingested*
from the .sav files (`defun_nofetal_raw`), this notebook looks at the
*documented* metadata, i.e. the DDI codebooks (`ddi_defun_nofetal_vars`),
for Defunciones No Fetales (non-fetal/general deaths). The DDI is what
tells us how a variable is supposed to be encoded (its measurement type,
its label, and the value/label pairs for its categories), independently of
whatever ended up in the actual data file for a given year. Here we want
to answer:

* which variables does DANE actually document, and in which years
* does a variable's measurement type (`intrvl`) or label text stay stable
  over time
* how many categories does each variable have per year, and does the DDI
  actually include a label for each of them (some years/variables only
  document the codes, not the labels -- this is what broke DDI ingestion for
  2023/2024, see `ingest_eevv_ddi.R`)
* for variables with a manageable number of categories, which codes are
  added/dropped over time, and does the label text for the same code drift
  (spelling, accents, rewording) across years

```{r}
ddi_summary <- ddi_defun_nofetal_vars |>
  mutate(
    n_categories = purrr::map_int(catgry, nrow),
    n_missing_label = purrr::map_int(catgry, \(x) sum(is.na(x$labl))),
    pct_missing_label = dplyr::if_else(
      n_categories > 0, n_missing_label / n_categories, NA_real_
    )
  )
```

# Coverage: variables documented per year

```{r}
ddi_summary |>
  group_by(year) |>
  summarize(
    n_vars = n(),
    n_vars_with_categories = sum(n_categories > 0),
    .groups = "drop"
  ) |>
  tidyr::pivot_longer(starts_with("n_vars"), names_to = "metric") |>
  ggplot(aes(x = year, y = value, color = metric, group = metric)) +
  geom_line() +
  geom_point() +
  theme(axis.text.x = element_text(angle = -45, hjust = 0), legend.position = "bottom")
```

The DDI does not document the same variables every year (some variables
were only introduced, renamed or dropped at some point -- e.g. this dataset
carries the cause-of-death variables, which are a whole subject of their
own). The next heatmap shows, for every variable/year combination, how many
categories the DDI documents for it (darker/lighter = more/fewer
categories; grey = the variable simply isn't documented that year).

```{r fig.height=14}
plotly::ggplotly(
  ddi_summary |>
    ggplot(aes(
      x = year, y = name, fill = n_categories,
      text = glue::glue(
        "\n",
        "intrvl: {intrvl}",
        "label: {labl}",
        "n_categories: {n_categories}",
        "n_missing_label: {n_missing_label}",
        .sep = "\n"
      )
    )) +
    geom_tile() +
    scale_fill_viridis_c(trans = "log1p", na.value = "grey90")
) |>
  plotly::layout(
    legend = list(orientation = "h"), xaxis = list(title = NA, tickangle = -45)
  )
```

# Measurement type (`intrvl`) over time

```{r fig.height=14}
plotly::ggplotly(
  ddi_summary |>
    ggplot(aes(
      x = year, y = name, fill = intrvl,
      text = glue::glue(
        "\n",
        "intrvl: {intrvl}",
        "label: {labl}",
        .sep = "\n"
      )
    )) +
    geom_tile()
) |>
  plotly::layout(
    legend = list(orientation = "h"), xaxis = list(title = NA, tickangle = -45)
  )
```

Variables whose `intrvl` is not the same across every year they appear in
(worth a manual look, a future `homologate_vars_defun_nofetal()` may need
to cast them explicitly):

```{r}
ddi_summary |>
  group_by(name) |>
  summarize(
    n_types = n_distinct(intrvl, na.rm = TRUE),
    types = paste(sort(unique(intrvl)), collapse = ", "),
    years = paste(sort(unique(year)), collapse = ", "),
    .groups = "drop"
  ) |>
  filter(n_types > 1) |>
  DT::datatable()
```

# Variable label (`labl`) consistency over time

The variable label is free text, so it is expected to drift a bit (accents,
capitalization). Still, worth a look to catch actual renames (a variable
being repurposed) as opposed to just cosmetic differences. This is the same
kind of check `get_var_labels_only()` already does to pick one label per
variable; here we show the years in which each wording was used.

```{r}
label_changes <- ddi_summary |>
  group_by(name) |>
  summarize(n_labels = n_distinct(labl), .groups = "drop") |>
  filter(n_labels > 1)

ddi_summary |>
  semi_join(label_changes, by = "name") |>
  group_by(name, labl) |>
  summarize(years = paste(sort(unique(year)), collapse = ", "), .groups = "drop") |>
  arrange(name) |>
  DT::datatable()
```

# Missing category labels

Some years document the category *codes* for a variable but not their
*labels* (`catValu` without a `labl`) -- this is what made DDI ingestion of
2023/2024 fail in the first place (see `ingest_eevv_ddi.R` /
`daner::get_var_categories()`), and it is worth monitoring going forward,
since it silently limits what `encode_ddi_factor()` can do with that
year/variable. The heatmap below shows the share of categories missing a
label (0% = fully labeled, white = no categories documented at all).

```{r fig.height=14}
plotly::ggplotly(
  ddi_summary |>
    filter(n_categories > 0) |>
    ggplot(aes(
      x = year, y = name, fill = pct_missing_label,
      text = glue::glue(
        "\n",
        "n_categories: {n_categories}",
        "n_missing_label: {n_missing_label}",
        .sep = "\n"
      )
    )) +
    geom_tile() +
    scale_fill_viridis_c(limits = c(0, 1), labels = scales::percent)
) |>
  plotly::layout(
    legend = list(orientation = "h"), xaxis = list(title = NA, tickangle = -45)
  )
```

# Category codes over time (small code lists)

Large code lists (cause-of-death codes and the like) are impractical to
inspect one tile at a time, so this section focuses on variables with a
manageable number of categories (at most 15, across all years combined),
which is where genuinely interesting encoding changes tend to happen (e.g. a
new category introduced, an old one retired, a "sin información" code
appearing/disappearing).

```{r results="asis"}
small_vars <- ddi_summary |>
  group_by(name) |>
  summarize(max_n = max(n_categories), .groups = "drop") |>
  filter(max_n > 0, max_n <= 15) |>
  arrange(name) |>
  pull(name)

catgry_small <- ddi_defun_nofetal_vars |>
  filter(name %in% small_vars) |>
  select(name, year, catgry) |>
  tidyr::unnest(catgry)

purrr::walk(small_vars, function(v) {
  cat("## ", v, "\n\n")

  p <- catgry_small |>
    filter(name == v) |>
    ggplot(aes(
      x = year, y = catValu, fill = catValu,
      text = glue::glue("labl: {labl}")
    )) +
    geom_tile() +
    theme(legend.position = "none")

  print(htmltools::tagList(
    plotly::ggplotly(p) |>
      plotly::layout(xaxis = list(title = NA, tickangle = -45))
  ))

  cat("\n\n")
})
```

# Category label drift for persisting codes

For a code (`catValu`) that shows up in more than one year for the same
variable, does its label text actually stay the same? Small wording/accent
differences are expected and harmless; a code being re-labelled to mean
something substantively different is the thing worth catching here.

```{r}
ddi_defun_nofetal_vars |>
  select(name, year, catgry) |>
  tidyr::unnest(catgry) |>
  filter(!is.na(labl)) |>
  group_by(name, catValu) |>
  summarize(
    n_distinct_labl = n_distinct(labl),
    labels = paste(unique(labl), collapse = " | "),
    years = paste(sort(unique(year)), collapse = ", "),
    .groups = "drop"
  ) |>
  filter(n_distinct_labl > 1) |>
  arrange(name, catValu) |>
  DT::datatable()
```
