---
title: "Inspect vars info over time"
output: 
  html_notebook:
    code_folding: hide
---

```{r}
vars_info <- tar_read(vars_info)
```

```{r results="asis"}
nac_raw <- tar_read(nac_raw)
```


To quickly visuallize what has changed, we'll take
a look at a few plots that show i) which variables exist in which years 
and ii) how are those variables encoded (class, type and unique values). 
This is an input to individually decide what to do with each variable 
(e.g. include/exclude, how to homologate them when categories have changed).
And it should help later down the line, when we -if and when- want to update
this analysis with newly published data (simply add the data and take a look 
at the last year to see if there is something different).

# Data as Ingested

```{r}
plotly::ggplotly(
  vars_info |>
    mutate(value = 1) |>
    ggplot(aes(x = year, y = var_names, fill = value, text = glue::glue(
      "\n",
      "var_class: {var_class}",
      "var_type: {var_type}",
      "var_unique: {var_unique}",
      .sep = "\n"
    ))) +
    geom_tile() +
    theme(legend.position = "none")
) |>
  plotly::layout(
    legend = list(orientation = "h"), xaxis = list(title = NA, tickangle = -45)
  )
```

```{r}
plotly::ggplotly(
  vars_info |>
    ggplot(aes(x = year, y = var_names, fill = var_class, text = glue::glue(
      "\n",
      "var_class: {var_class}",
      "var_type: {var_type}",
      "var_unique: {var_unique}",
      .sep = "\n"
    ))) +
    geom_tile()
) |>
  plotly::layout(
    legend = list(orientation = "h"), xaxis = list(title = NA, tickangle = -45)
  )
```

```{r}
plotly::ggplotly(
  vars_info |>
    ggplot(aes(x = year, y = var_names, fill = var_type, text = glue::glue(
      "\n",
      "var_class: {var_class}",
      "var_type: {var_type}",
      "var_unique: {var_unique}",
      .sep = "\n"
    ))) +
    geom_tile()
) |>
  plotly::layout(
    legend = list(orientation = "h"), xaxis = list(title = NA, tickangle = -45)
  )
```

```{r}
plotly::ggplotly(
  vars_info |>
    ggplot(aes(x = year, y = var_names, fill = var_unique, text = glue::glue(
      "\n",
      "var_class: {var_class}",
      "var_type: {var_type}",
      "var_unique: {var_unique}",
      .sep = "\n"
    ))) +
    geom_tile() +
    theme(legend.position = "none")
) |>
  plotly::layout(
    legend = list(orientation = "h"), xaxis = list(title = NA, tickangle = -45)
  )
```

# DDI Ingested Categories

```{r}
ddi_nac_vars <- tar_read(ddi_nac_vars)

ddi_nac_vars <- ddi_nac_vars |>
  mutate(catgry = purrr::map_chr(catgry, \(x)  {
    paste0("\n", paste0(paste(x$catValu, x$labl, sep = ": "), collapse = "\n"))
  }))
```

```{r}
plotly::ggplotly(
  ddi_nac_vars |>
    ggplot(aes(x = year, y = name, fill = catgry, text = glue::glue(
      "\n",
      "intrvl: {intrvl}",
      .sep = "\n"
    ))) +
    geom_tile() +
    theme(legend.position = "none")
) |>
  plotly::layout(
    legend = list(orientation = "h"), xaxis = list(title = NA, tickangle = -45)
  )
```


# Checking vars

```{r}
already_checked <- c()
```

```{r results="asis"}
being_checked <- c("ano")
already_checked <- c(already_checked, being_checked)
cat("## ", being_checked)
```

it's all good, data available for all years (surprise!), only
thing to fix: some years are encoded as character and some as numeric
so, let's make them all character (or factor?), 'cause anyway this
variable will mostly (always?) be used as a categorical variable'

```{r results="asis"}
being_checked <- c("apgar1", "apgar2")
already_checked <- c(already_checked, being_checked)
cat("## ", being_checked)
```

until 2007 grouped values as follows

- 1: apgar <5
- 2: apgar between 5-6
- 3: apgar >=7
- 9: sin información

from 2008 on, integers between 1 and 10, and 99 for "sin información",
and some observations with empty values (so, not explicitly 99, nor NA, but "").
Also, some years encoded as numeric, and some as character, with values e.g. 01
so we want to homologate this as follows:

* create apgar1_old and apgar2_old, recoding the periods from 2008, to use
  the groupings used in the years before 2008
* keep apgar1 and apgar2 from 2008 on, and set the previous as NA
* Convert "sin información", and empty values, to NA

```{r results="asis"}
being_checked <- c("area_res", "areanac")
already_checked <- c(already_checked, being_checked)
cat("## ", being_checked)
```

- all years share the values 1, 2, 3, and the 9 for missing value, just not 
  present in some of the latest years. Some years have empty values ""
- catgry, pretty consistent for both variables. Only 2018 does not have category
  labels (bundled with 2017, so perhaps because of this)

* just cast everything to numeric
* make 9 to missing values
* and finally encode with the DDI categories of any of the years (except 2018)

```{r results="asis"}
being_checked <- c("aten_par")
already_checked <- c(already_checked, being_checked)
cat("## ", being_checked)
```

- all years share the values 1-6, plus 9 for missing value, just not 
  present in some of the latest years. 
- catgry, pretty consistent for all years. Only 2018 does not have category
  labels (bundled with 2017, so perhaps because of this)

* just cast everything to numeric
* make 9 to missing values
* and finally encode with the DDI categories of any of the years (except 2018)


```{r results="asis"}
being_checked <- c(
  "cod_dpto", "cod_inst", "cod_munic", "codmunre", "codpres", "codptore"
)
already_checked <- c(already_checked, being_checked)
cat("## ", being_checked)
```

```{r results="asis"}
nac_raw |>
  filter(year == 2013) |>
  pull(sav_data) |>
  first() |>
  pull(codptore) |>
  glimpse() |>
  attributes()
```

- cod_dpto, some years come as haven_labelled, but we would rather just keep
  it as a string, hence, just DANE's code to two digits
- cod_inst, no categories metadata, just the variable lable. Only 
  available until 2017. TODO: check what happened there
- cod_munic, no categories metadata, just the variable lable.
- codmunre, metadata shows 999 as sin información. Should we explicitly make it 
  missing (NA)?, or leave that for later. TODO: decide about missing
- codpres, metadata shows 170 colombia, 999 sin información. DDI the same categories
  until 2003, 2004 introduced 998 resto de países, but from 2008 to 2013 no DDI
  categories and from 2014 on, only 170 colombia and 999 sin información
- codptore, just a few years with metadata. 01, sin información

* cast them all to character. 
* Yet, a few years are originally encoded as numeric thus, it may require 
  further processing down the line, for example, via daner to make valid 
  2 or 3 digits codes

```{r results="asis"}
being_checked <- c("edad_madre", "edad_padre")
already_checked <- c(already_checked, being_checked)
cat("## ", being_checked)
```

```{r results="asis"}
nac_raw |>
  filter(year == 2015) |>
  pull(sav_data) |>
  first() |>
  pull(edad_padre) |>
  glimpse()
```

- all years available
- edad_madre has consistent unique values and DDI categories across all years
- edad_padre is weird, do not show unique values. Let's investigate: DONE:
  edad_padre is expressed in years (99 used for sin información), 
  while edad_madre contains categories for age brackets

* encode edad_madre using DDI categories for any year (they are all the same)
* keep edad_padre as numeric, and just conver 99 -> NA


```{r results="asis"}
being_checked <- c("est_civm")
already_checked <- c(already_checked, being_checked)
cat("## ", being_checked)
```

```{r results="asis"}
ddi_nac_vars |>
  filter(year == 2015, name == "est_civm") |>
  pull(catgry) |>
  first()
```

- available all years, consistent values, a few years with metadata.
- DDI shows that the categories changed from 2008 on, so need to homologate

__Until 2007__
1       Soltera
2       Casada
3       Viuda
4       En unión libre
5       Separada o divorciada
9       Sin información

__From 2008 on__
1       No está casada y lleva dos o más años viviendo con su pareja
2       No está casada y lleva menos de dos años viviendo con su pareja
3       Está separada, divorciada
4       Está viuda
5       Está soltera
6       Está casada
9       Sin información

* keep est_civim encoding it with DDI categories, from 2008 on (and NA before that)
* homologate >=2008 into <= 2007 as follows:
  6 -> 2
  5 -> 1
  4 -> 3
  3 -> 5
  2 -> 1
  1 -> 4



```{r results="asis"}
being_checked <- c("gru_san")
already_checked <- c(already_checked, being_checked)
cat("## ", being_checked)
```

- only available until 2007
- but otherwise pretty consistent, with complete DDI categories

* only chnage 9 to NA and encode with DDI


```{r results="asis"}
being_checked <- c("idclasadmi")
already_checked <- c(already_checked, being_checked)
cat("## ", being_checked)
```

- only available from 2008
- but otherwise relatively consistent, same values, with just some years with 
  empty "" and chnagins data type but ok for underlying values
- with complete DDI categories, only missing for 2018

* only chnage 9 and implicitly empty (i.e. "") to NA and encode with DDI


```{r results="asis"}
being_checked <- c("idclasadmi")
already_checked <- c(already_checked, being_checked)
cat("## ", being_checked)
```

- only available from 2008
- but otherwise relatively consistent, same values, with just some years with 
  empty "" and chnagins data type but ok for underlying values
- with complete DDI categories, only missing for 2018

* only chnage 9 and implicitly empty (i.e. "") to NA and encode with DDI


```{r results="asis"}
being_checked <- c("idfactorrh", "idhemoclas")
already_checked <- c(already_checked, being_checked)
cat("## ", being_checked)
```

- only available from 2008
- but otherwise very consistent
- with complete DDI categories, only missing for 2018

* only chnage 9 and implicitly empty (i.e. "") to NA and encode with DDI

* and at this point, create new vars, homologating gru_san into the post 2007
  vars idfactorrh and idhemoclas. Here's the unique values of those columns, 
  so you basically need to extract what is within the parenthesis from column gru_san
  and separate that into factorrh and hemoclas
* idfactorrh: "Negativo" "Positivo"
* idhemoclas: "O"  "B"  "A"  "AB"
* gru_san: "O+ (O positivo)"   "O- (O negativo)"   "AB+ (AB positivo)" 
  "A+ (A positivo)"   "A- (A negativo)"   "B+ (B positivo)"   "AB- (AB negativo)" 
  "B- (B negativo)" 


```{r results="asis"}
being_checked <- c("idpertet", "idpuebloin")
already_checked <- c(already_checked, being_checked)
cat("## ", being_checked)
```

- idpertet available from 2008 until 2016, values and DDI consistent
- idpertet available from 2017 on, DDI consistent except for missing
  2018, and values consistent, except that the numeric encoded years do
  not have the 9 missing value (just missing?)

* make 9 to NAs and encode each with DDI


```{r results="asis"}
being_checked <- c("mes")
already_checked <- c(already_checked, being_checked)
cat("## ", being_checked)
```

- values and DDI ok, but let's keep just the number

* make it numeric (or factor?)

```{r results="asis"}
being_checked <- c("mul_parto")
already_checked <- c(already_checked, being_checked)
cat("## ", being_checked)
```

- All values and DDI consistent

* 9 -> NA and encode DDI

```{r results="asis"}
being_checked <- c("n_emb", "n_hijosv")
already_checked <- c(already_checked, being_checked)
cat("## ", being_checked)
```

- Values seem ok, DDI only indicidate 99 as sin información

* 99 -> NA and keep it as numeric


```{r results="asis"}
being_checked <- c("niv_edum", "niv_edup")
already_checked <- c(already_checked, being_checked)
cat("## ", being_checked)
```

- categories changed in 2008, but otherwise values and DDI consistent

Categories pre 2008
    1       Preescolar
    2       Primaria completa
    3       Primaria incompleta
    4       Secundaria completa
    5       Secundaria incompleta
    6       Universitaria completa
    7       Universitaria incompleta
    8       Ninguno
    9       Sin información

Categories from 2008 on
    01      Preescolar
    02      Básica primaria
    03      Básica secundaria        
    04      Media académica o clásica
    05      Media técnica
    06      Normalista
    07      Técnica profesional
    08      Tecnológica
    09      Profesional
    10      Especialización
    11      Maestría
    12      Doctorado
    13      Ninguno
    99      Sin información


* rename to old and new, to keep both encodings separated
* TODO: perhaps homologate to get a whole?

```{r results="asis"}
being_checked <- c("nom_inst", "nomclasad", "otro_sit", "otrparatx")
already_checked <- c(already_checked, being_checked)
cat("## ", being_checked)
```

- these are mostly character vectors with many values

* just keep them as character


```{r results="asis"}
being_checked <- c("numconsul")
already_checked <- c(already_checked, being_checked)
cat("## ", being_checked)
```

- consistent values, 99 sin información

* keep it numeric and 99 -> NA

```{r results="asis"}
being_checked <- c("peso_nac")
already_checked <- c(already_checked, being_checked)
cat("## ", being_checked)
```

- consistent values and DDI categories throughout

* 9 -> NA and encode with DDI


```{r results="asis"}
being_checked <- c("profesion")
already_checked <- c(already_checked, being_checked)
cat("## ", being_checked)
```

- categories changed in 2008
- but otherwise values and DDI categories are consistent throughout

* create profesion_old and profesion_new, replacein 9 -> NA and encoding DDI
* TODO: should we create a profesion_homo?, perhaps merging enfermería?


```{r results="asis"}
being_checked <- c("seg_social")
already_checked <- c(already_checked, being_checked)
cat("## ", being_checked)
```

- categories changed in 2008
- but otherwise values and DDI categories are consistent throughout

* create _old and _new, replacein 9 -> NA and encoding DDI
* TODO: should we also change "6: Ignorado" to NA, in the pre-2008 data?
* TODO: should we create a _homo?

```{r results="asis"}
being_checked <- c("sexo")
already_checked <- c(already_checked, being_checked)
cat("## ", being_checked)
```

- categories changed in 2014, introducing sexo: indeterminado
- but otherwise values and DDI categories are consistent throughout

* keep only variable sexo, but encode all of them with 2014 DDI


```{r results="asis"}
being_checked <- c("sit_parto")
already_checked <- c(already_checked, being_checked)
cat("## ", being_checked)
```

- all values and DDI categories are consistent throughout

* just 9 -> NA and encode DDI


```{r results="asis"}
being_checked <- c("t_ges")
already_checked <- c(already_checked, being_checked)
cat("## ", being_checked)
```

- all values and DDI categories are consistent throughout

* just 9 -> NA and encode DDI

```{r results="asis"}
being_checked <- c("talla_nac")
already_checked <- c(already_checked, being_checked)
cat("## ", being_checked)
```

- all values and DDI categories are consistent throughout

* just 9 -> NA and encode DDI


```{r results="asis"}
being_checked <- c("tipo_parto")
already_checked <- c(already_checked, being_checked)
cat("## ", being_checked)
```

- all values and DDI categories are consistent throughout

* just 9 -> NA and encode DDI


```{r results="asis"}
being_checked <- c("ultcurmad", "ultcurpad")
already_checked <- c(already_checked, being_checked)
cat("## ", being_checked)
```

- DDI only has category 99: sin información

* make it a numeric and replace 99 -> NA


## Vars pending screening
```{r}
vars_info$var_names |>
  unique() |>
  setdiff(already_checked) |>
  sort()
```



# TODO 
do date later

```{r results="asis"}
being_checked <- c("fecha_nacm")
already_checked <- c(already_checked, being_checked)
cat("## ", being_checked)
```

```{r results="asis"}
nac_raw |>
  filter(year == 2019) |>
  pull(sav_data) |>
  first() |>
  pull(fecha_nacm) |>
  glimpse() |>
  identity()
```




